---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
llm_path <- Sys.getenv(
  "RLLM_README_GGUF",
  "LFM2.5-8B-A1B-Q4_K_M.gguf"
)
llm_available <- file.exists(llm_path)
```

# Rllm

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Rllm is the **composition layer** of the
[Rfmalloc](https://github.com/sounkou-bioinfo/Rfmalloc) out-of-core array
stack: it wires [Rfmalloc](../Rfmalloc) (file-backed, memory-mapped storage),
[Rgguf](../Rgguf) (GGUF model files as typed storage views) and
[Rggml](../Rggml) (vendored GGML compute with runtime-SIMD-dispatched
quantized kernels) into LLM
inference - quantized weights stay in their GGUF encoding, memory-mapped from
disk, and are never decoded to double on the compute path.

The first composition is a codec-aware matrix-product backend:
`dense %*% quantized_fmalloc_tensor` contracts the encoded blocks directly
through GGML after registering with Rfmalloc's backend registry. The second
normalizes GGUF metadata into one serializable semantic program, binds that
program to mapped weights and lowers it into forward graphs, state, embeddings
and byte-level generation over the same typed storage. Llama, Qwen3.5,
LFM2MoE and EmbeddingGemma are model probes of the operator vocabulary rather
than C++ subclasses. Their metadata adapters trace the program and declare its
typed parameters directly. `rllm_plan()` constructs a layer-oriented view only
when inspection is requested.

## Quantized products, zero-copy

```{r bridge}
library(Rllm)   # registers + selects the ggml backend

rt <- Rfmalloc::open_fmalloc(tempfile(fileext = ".bin"))
set.seed(1)
W  <- matrix(rnorm(512 * 8, sd = 0.4), nrow = 512)
Wq <- rllm_quantize_tensor(W, "q4_k", runtime = rt)  # 4.5 bits/weight, mmap'd
X  <- matrix(rnorm(4 * 512), nrow = 4)
Y  <- X %*% Wq                                       # GGML SIMD, no f64 decode
sqrt(sum((Y - X %*% W)^2) / sum((X %*% W)^2))        # q4_k-accurate
```

## Inference: bytes in, bytes out

The model I/O boundary is **token ids and raw bytes** - never character
vectors in the core. `rllm_encode()`/`rllm_decode()` are edge codecs built
from the GGUF's own byte-level BPE metadata (no external tokenizer);
`rawToChar()` is the caller's interpretation. The same boundary serves
byte-level, vision, or pure-codec models later.

```{r infer, eval = llm_available, cache = TRUE, cache.path = "README_cache/llm/", cache.extra = if (llm_available) unname(file.info(llm_path)[c("size", "mtime")])}
local({
  backing <- tempfile(fileext = ".bin")
  rt <- Rfmalloc::open_fmalloc(backing, mode = "scratch", size_gb = 2)
  on.exit({
    Rfmalloc::cleanup_fmalloc(rt)
    unlink(backing)
  }, add = TRUE)

  model_path <- Sys.getenv(
    "RLLM_README_GGUF",
    "LFM2.5-8B-A1B-Q4_K_M.gguf"
  )
  model <- rllm_gguf_model(model_path, runtime = rt)
  gen <- rllm_generate(
    model,
    charToRaw("The capital of France is"),
    n_new = 16L,
    runtime = rt
  )
  rawToChar(gen$raw)
})
```

The evaluated transcript comes from LFM2.5-8B-A1B `Q4_K_M`. Set
`RLLM_README_GGUF` when rendering the Rmd. An absent model skips evaluation;
the document does not substitute a toy model or recorded success text. The
knitr cache is keyed by the model file's size and modification time.

## Architecture programs are R

An architecture is described with ordinary R functions. `rllm_module()` gives
a `forward` function a persistent identity, the base pipe records dataflow,
`rllm_residual()` records branches, `rllm_tap()` names intermediate outputs and
`rllm_loop()` preserves shared recurrence instead of unrolling it. Freezing the
trace with `rllm_program()` removes the builder environment and leaves only
serializable data. `rllm_inputs()` starts a trace with several typed inputs,
while an operator's `outputs` specification keeps several named results in the
same graph. Masks, RoPE, routing, scale, state and tensor parameters remain
direct named node attributes rather than being hidden in an architecture blob.

```{r program}
d <- "hidden"
norm <- rllm_parameter("encoder.norm.weight", d)

encoder_block <- rllm_module("encoder.block", function(x) {
  x |>
    rllm_residual(function(branch) {
      branch |>
        rllm_norm(norm) |>
        rllm_op(
          "attention",
          mask = list(type = "bidirectional")
        )
    })
})

program <- rllm_input(
  "embedding", c(feature = d, sequence = "n_token")
) |>
  encoder_block() |>
  rllm_tap("per_token") |>
  rllm_pool("mean") |>
  rllm_program("encoder")

program
```

The Qwen3.5 program is the first deliberately heterogeneous model probe. The
metadata and tensor directory of
[Ternary Bonsai 27B](https://huggingface.co/prism-ml/Ternary-Bonsai-27B-gguf)
normalize to one R data
AST with 48 gated-delta recurrent blocks, 16 gated-attention blocks, two kinds
of persistent state and 851 typed parameter bindings. Its official group-64
`q2_0` weights remain a storage-codec concern. The recurrent and attention
mixture remains an architecture concern. Neither becomes a model-family
branch in C++. Both operators lower through Rggml, including MRoPE,
head-interleaved query gates, causal convolution and recurrent matrix state.
Whole-batch and incremental execution are pinned to a pure-R equation oracle.
For the real checkpoint, a token-zero forward pass produces all 248,320 logits
bit-for-bit identically to upstream llama.cpp built with the corresponding
portable CPU backend. The pinned GGML release has no `q2_0` CUDA or Vulkan
matrix-product kernel, so this ternary checkpoint executes on CPU. The same
Qwen3.5 graph can use CUDA when its weights use a codec that backend supports.

The same frozen program has two consumers. `rllm_execute()` interprets dataflow
and structured control flow through an explicit table of dense R operators.
Its reference vocabulary pins Tiny Recursive Model evaluation against a direct
R recurrence: `z_H` and `z_L` cross three nested symbolic loops while one
reasoning module stays shared. `rllm_forward()` binds the program's parameters
to their mapped GGUF spans, validates the supported transformer grammar once
and lowers its operators through GGML. There is no separate native model plan.
The full attention and SwiGLU TRM topology is serializable, but a Samsung
checkpoint importer and native bidirectional-attention lowering do not yet
exist.

The native lowerer implements the operators exercised by llama, Qwen3.5,
LFM2MoE and EmbeddingGemma. ESM-2 8M also crosses the real numerical boundary.
`tools/convert_esm2.R` reads Facebook's official `model.safetensors`,
`config.json` and `vocab.txt`, maps their published Transformers names back to
the fair-esm state-dict names, and gives the 106 unmodified F32 tensors to
Rgguf's upstream GGUF writer. From the package source,
`Rscript tools/convert_esm2.R esm2.gguf` downloads those three official files;
an optional second argument selects a cached directory or another base URL.
The GGUF adapter constructs
a 69-node program with typed token and padding inputs, six rotary attention
layers returning hidden states and attention maps, tied embeddings,
representation and attention taps, and the contact head. The dense R
interpreter matches the official execution on a fixed protein: selected
logits, representations, attention probabilities and contacts agree within
0.003, 0.0002, 0.00003 and 0.0003 respectively. Its native two-input GGML
lowering remains explicit work and fails at the grammar boundary.

[OpenSpliceAI](https://github.com/Kuanhao-Chao/OpenSpliceAI) supplies a second
non-transformer numerical test. Its program records the residual dilated 1-D
convolutions, inference batch normalization, leaky ReLU, accumulated skip
projections, context crop and per-position softmax without an OpenSpliceAI
executor branch. `tools/convert_openspliceai.R` uses a Python environment with
PyTorch and safetensors only to read the upstream state dictionary into a
temporary safetensors file, then gives all 56 F32 tensors to Rgguf's writer:
`Rscript tools/convert_openspliceai.R openspliceai.gguf model_80nt_rs10.pt 80`.
The complete 16-position output of the MANE 80 nt rs10 checkpoint agrees with
upstream PyTorch within 5e-7. The generic GGML transformer lowerer rejects its
floating-point sequence-input grammar explicitly.

Evo 2 remains an executable-frontier probe, but its program is not a
caricature. The Evo 2 7B program records the complete 32-layer schedule: 27
short, medium and long Hyena cascades with their distinct FIR or IIR state,
interleaved with five causal-attention layers. Its missing checkpoint importer
and numerical lowering fail explicitly; they are not model-family branches
hidden in C++.

[Rtinycc](https://github.com/sounkou-bioinfo/Rtinycc) is a plausible prototype
target for generated C reference operators and ABI glue. It can compile a
lowering in memory and test it against the R semantics before hot operators are
promoted into Rggml. It is not treated as a substitute for GGML's tuned tensor
kernels.

A recorded run with SmolLM2-135M `Q4_K_M` (30 layers, grouped-query
attention, a `q5_0`-dominant quantization mix) used a 12-token prompt and 128
greedy decode tokens. The median of three complete generations was 40.2
tokens/s on CPU and 69.7 tokens/s on the RTX 5050 CUDA path after its one-time
weight upload.

`rllm_gguf_model()` reads the hyperparameters, constructs the semantic program
and borrows every weight's exact read-only span. Binding validates every
declared parameter name and shape before native execution. No weight is copied
into the Rfmalloc work file. On CPU, `rllm_forward()` points GGML tensors at
those bytes while assembling RMSNorm, RoPE, causal grouped-query attention and
SwiGLU through Rggml's C-callables; Rllm links no private GGML copy.

`rllm_kv_cache()` allocates one key and value slab per layer in plain R memory
or in fmalloc storage. The latter makes the cache file-backed, evictable and
open to a different cache codec without changing the graph. `rllm_generate()`
prefills that cache and then performs EOS-aware decoding one token at a time.

## Correctness discipline

The forward pass is pinned to pure-R implementations of the same arithmetic
for llama, Qwen3.5, LFM2MoE and EmbeddingGemma. Hermetic f32 models require
incremental state to equal whole-batch execution at every position. Qwen3.5 is
also compared over every real-model logit with an upstream build using the same
portable CPU arithmetic. On a real quantized MoE, changing batch width can
select a different GGML GEMV or GEMM kernel; tiny router-score rounding may
cross the top-k boundary, so the real-file test pins the chosen token and logit
direction while generation is checked against the upstream continuation.
Every codec decoder must be bit-identical to GGML's type-traits `to_float`, the
cross-validation that caught the Q4_K bug in the earlier partial C port. At
the byte boundary,
`rllm_decode(rllm_encode(x))` must always equal `x`.

The hermetic tests write their synthetic GGUF models at test time with
[Rgguf](../Rgguf)'s own writer. `RLLM_TEST_GGUF=<path>` exercises a real
generation model, while `RLLM_EMBEDDING_GGUF=<path>` compares the 300M
EmbeddingGemma Q8_0 probe with an upstream llama.cpp reference.
`RLLM_QWEN35_GGUF=<path>` executes a real hybrid checkpoint; pairing it with
`RLLM_QWEN35_REFERENCE=<f32.bin>` checks a portable upstream logit transcript
bit-for-bit. `RLLM_ESM2_GGUF=<path>` and
`RLLM_OPENSPLICEAI_GGUF=<path>` run the two real biological-model numerical
oracles.

## Compute follows residency

The CPU path borrows the GGUF mapping directly and uses GGML's
runtime-dispatched quantized kernels: AVX2 selected by CPUID on x86 and the
mandatory NEON baseline on aarch64, with no non-portable flags recorded in the
package. Dense products remain available to R's BLAS.

The optional CUDA path executes the same graph. Its model-owned context uploads
the codec-native weights once and retains them in device memory; subsequent
forward passes move only mutable inputs, cache state and logits through
Rggml's backend-neutral transfer interface. The host GGUF mapping remains the
authoritative storage, while device residency is an execution decision rather
than another model format. Vulkan also ships in Rggml, but Rllm does not select
it. The rig suite checks whole-batch and cached CUDA logits, plain and
fmalloc cache slabs, and cache handoff in both directions between CPU and CUDA.

The suite distinguishes cache correctness from invariance to batch shape.
For an identical prefix shape, cached and uncached logits from the real
SmolLM2 `Q4_K_M` file are bit-identical on both CPU and CUDA. CPU also gives
the same result when a four-token prefix is evaluated whole or incrementally.
CUDA does not: on the structural token sequence 1, 5, 9, 2,
prefix-versus-whole cosine similarity ranges from 0.761 to 0.997 and two of
four argmaxes differ. Real-shape products for all four quantized weight types
remain within 1.25e-4 NMSE of CPU, and CUDA one-column and four-column products
agree exactly. The remaining numerical question is therefore at graph level:
pin Rllm's result against the matching upstream GGML and llama.cpp paths for
each batch shape before claiming shape-invariant CUDA execution.

Persistent allocation by itself is not the missing acceleration. A measured
GGML scheduler experiment reduced CUDA decode from 69.7 to 63.7 tokens/s, and
CUDA graph capture reached 62.6 because the attention extent and graph
properties change at every token. Both experiments were removed from the
default path. Stable graph shapes and a device-resident mutable KV cache remain
open execution questions; the storage and GGUF interfaces need no fork for
either.

## Install

From the [r-universe](https://sounkou-bioinfo.r-universe.dev):

```{r install, eval = FALSE}
install.packages("Rllm",
  repos = c("https://sounkou-bioinfo.r-universe.dev", getOption("repos")))
```

or via a GitHub subdir ref: `pak::pak("sounkou-bioinfo/Rfmalloc/packages/Rllm")`.
The CPU stack targets Linux, macOS and Windows; the optional CUDA build is
Unix-like. GPL (>= 2).
