---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(Rpgen)
```

# Rpgen

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rfmalloc/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

Rpgen is the genomic record source for the Rfmalloc stack. It uses PLINK 2's
own pgenlib and import code to read PGEN, PLINK 1 BED, PED/MAP, TPED/TFAM,
VCF/BCF, BGEN, Oxford GEN, HAPS/legend, EIGENSTRAT, and legacy PLINK 1 dosage
data into file-backed compute layouts.

The central operation is `rpgen_ingest()`:

```{r ingest, message = FALSE}
library(Rfmalloc)

pgen <- system.file("extdata", "chr21_phase3_start.pgen", package = "Rpgen")
backing <- tempfile(fileext = ".bin")
rt <- open_fmalloc(backing, mode = "scratch", size_gb = 1)

dose <- rpgen_ingest(
  pgen,
  representation = "dosage",
  runtime = rt
)

cat(sprintf(
  "%d samples x %d variants, %d payload bytes\n",
  nrow(dose), ncol(dose), length(unclass(dose))
))
```

```{r ingest_cleanup, include = FALSE}
cleanup_fmalloc(rt)
unlink(backing)
```

## The abstraction

A format reader emits bounded panels of semantic records. Rfmalloc owns the
destination allocation and layout:

| Representation | Meaning | Storage |
|---|---|---|
| `"hardcall"` | `0`, `1`, `2`, or missing non-reference count | 2 bits per sample and variant |
| `"dosage"` | fractional non-reference count | 1-byte fixed point |
| `"haplotype"` | two phased ref/non-ref bits per sample | locus-major, 1 bit per call |
| `"f64"` | full-precision dosage | file-backed doubles |

This is the useful split. A source declares whether it produced integers,
doubles, or packed phase bits. The destination decides whether those records
remain full precision or become a compute codec. Readers do not know codec
headers, alignment, or allocation details.

PGEN and BED keep one pgenlib reader open and transfer bounded variant panels
directly into that destination. They never construct a complete genotype
matrix. Haplotype reads use `PgrGetP()` and preserve phase. A missing genotype
or unphased heterozygote is an error because the one-bit HMM layout cannot
represent either without a separate mask.

## Format closure at the record sink

| Source | Companion arguments | Current route |
|---|---|---|
| PGEN | optional `pvar` | persistent pgenlib reader |
| PLINK 1 BED | optional `bim`, `fam` | persistent pgenlib reader |
| PLINK 1 PED/MAP | `map` | importer, bounded transpose scratch, record sink |
| PLINK 1 TPED/TFAM | `tfam` | importer to record sink |
| VCF/BCF | none | importer to record sink |
| BGEN | optional `sample` | importer to record sink |
| Oxford GEN | `sample` | importer to record sink |
| HAPS/legend | `legend`, `sample`, `chr` | bounded count, phased importer to record sink |
| EIGENSTRAT | `ind`, `snp` | importer to record sink |
| PLINK 1 dosage | `fam`, `map` | importer to record sink |

PLINK 2 already exposes each decoded variant immediately before
`STPgenWriter` encodes it. During `rpgen_ingest()`, Rpgen redirects every
writer append shape to Rfmalloc's record context, including multiallelic,
phase, dosage, and sparse records. No genotype PGEN is written or read back.
The importers may write temporary `.pvar`/`.psam` metadata companions, which
are removed on exit. PED/MAP also retains PLINK 2's sample-major transpose
scratch because changing that physical order is an actual algorithmic
transpose, not a package-boundary copy. HAPS needs a bounded preliminary line
count because its upstream writer receives an upper bound instead of the exact
variant count.

The standalone `rpgen_import_*()` functions do not activate this hook. They
continue to produce ordinary PGEN files for interoperability.

The dense `rpgen_read_hardcalls()` and `rpgen_read_dosages()` functions remain
available for small data, inspection, and reference comparisons. They are not
used by the bounded ingestion path.

## Phased HAPS

`rpgen_ingest(..., representation = "haplotype")` stores variants as aligned,
contiguous locus rows over donor haplotypes. This is the access direction used
by Li and Stephens HMM kernels such as kalis Forward/Backward:

```{r haps, eval = FALSE}
hap <- rpgen_ingest(
  "cohort.haps",
  format = "haps",
  representation = "haplotype",
  legend = "cohort.legend",
  sample = "cohort.sample",
  chr = "1",
  runtime = rt
)
```

## Provenance and license

The vendored pgenlib tree is generated by
`tools/vendor-pgenlib/vendorpgen.R`. The PLINK 2 importer closure is generated
by `tools/vendor-plink2-import/vendorplink2import.R`. Both recipes are pinned
and checked for drift in CI. See `inst/COPYRIGHTS` and the two `PROVENANCE.md`
files for source pins, local patches, and licenses.

Rpgen is GPL-3 because it combines with pgenlib's LGPL-3 code. The vendored
zstd, libdeflate, SIMDe, and SFMT components retain their own licenses.

## Installation

From the [r-universe](https://rgenomicsetl.r-universe.dev):

```{r eval = FALSE}
install.packages("Rpgen",
  repos = c("https://rgenomicsetl.r-universe.dev", getOption("repos")))
```

or from the monorepo subdirectory:

```{r eval = FALSE}
pak::pak("sounkou-bioinfo/Rfmalloc/packages/Rpgen")
```

Rpgen depends on Rfmalloc and builds on Linux, macOS, and Windows. CI also
checks Apple Silicon and Linux aarch64.
